Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-09 threshold/draft.Rnw
switched to bibtex, changed examples
Dropped rnaseq example and using a different small example

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$
be any two columns of the matrix
$A$, $1\le i,j\le n$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$--that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Let $t$ be a given correlation threshold, $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders the entries
of $v_1$ in increasing order.  For example, Figure \ref{fig1} displays the
ordered entries of $Pv_1$ for the small ``BicatYeast'' example (where
$A\in\mathcal{R}^{70\times 419}$) from Section \ref{examples}.  The lines in
the plot illustrate the interval associated with the correlation threshold
$t=0.95$ placed at an arbitrary vertical axis location.

\begin{figure}[!ht]
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(BicatYeast)
A=t(BicatYeast)
s=svd(A)
i=order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.95
abline(h=c(-0.05,-0.05+sqrt(2*(1-t))/s$d[1]),col=2,lwd=0.5)
@
\caption{
Example plot of the ordered entries of $v_1$ for the BicatYeast example from
Section \ref{examples}. The red lines illustrate the interval corresponding
threshold $t=0.95$ placed at an arbitrary location on the vertical axis. Points
farther apart than this interval correspond to column vectors that do not meet
the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

@@

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 49\% of
adjacent vectors fall below the correlation threshold of $0.95$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

A full/naive thresholded correlation computation requires $O(n^2 m)$ flops.
Let $p$ be the selected IRLBA dimension. Then the proposed method requires
$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in
$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
plus the flops required by IRLBA and the post-processing step 7 (also $O(np)$).

@@

We evaluated the algorithm using three public gene expression and methylation
datasets.  Most tests were performed on a desktop PC with a single AMD
A10-7850K 3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered
RAM running the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R
version 3.1.2 using double-precision arithmetic and the OpenBLAS library
version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}. The last experiment includes a test run on a Linux
cluster as described in the text.

@@

\subsection{Small BicatYeast example}

The first example is a tiny proof of concept that provides the data for Figure
\ref{fig1} and establishes our timing and memory use measurement methodology.
Wall-clock times are reported in seconds and peak memory use in excess of
memory required to store the input data matrix is reported in megabytes.

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
software\cite{biclust, bicat}. The data consist of 70 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 419
Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
example finds all pairs of columns (Affymetrix probes) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(irlba)
source("tcor.R")  # XXX Replace later with tcor pacakge
library(biclust)  # for the BicatYeast data
data(BicatYeast)  # in gene by sample orientation
A = t(BicatYeast) # to correlate genes to genes
threshold = 0.95

@@

# Fast brute force full correlation:
m2  = sum(gc()[,2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this tiny first example,
both in speed and slightly in memory use.  Each algorithm identifies the same
115 unique vector pairs that exceed the specified correlation threshold of
0.95, with comparative timing and memory use shown in the following table.

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "BicatYeast results, threshold=0.95", label = "BicatYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

\subsection{TCGA Gene Expression Examples}

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The next example uses RPKM-normalized gene expression data measured
across 20,532 genes from 878 breast cancer tumor samples obtained from the
Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the supplemental
code\cite{sup} (also embedded in this Sweave document) for examples of
efficiently converting the downloaded sample expression data into an R matrix.

Eight columns with constant zero values were removed from the data,
leaving an $878\times 20524$ matrix. The example computes
pairs of gene expression column vectors with correlation values of 0.99
or more. We use the same {\tt tcor} projection dimension of 10 and performance
measurement methodology as used in the first example.

Both methods find the same 3,572 pairs of vectors. The fast brute-force
computation is slightly faster than the {\tt tcor()} function largely due to
the still relatively small input matrix and better utilization of optimized
BLAS routines. But the {\tt tcor()} function uses far less memory.  (For
comparison, R's {\tt cor()} function took about 372 seconds and about
3200$\,$MB RAM on the same problem on the same test system.)

<<echo=FALSE, results=tex>>=
# --------------------------------------------------------------------------------------------------
# The first time you run this, you need to download the data as outlined in the
# following commented-out section. This code creates a file named
# 'rpkm.RData' used subsequently.
#
# To run any of the remaining (large) examples, first set the system environment
# variable RUN_EXAMPLES=TRUE (otherwise they won't run and representative results
# from Bryan's test machine will be printed).
#
#library(doMC)
# Download and extract this file with, for example:
# wget  http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# tar zxf gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# cd gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0
# R --slave < read.R   # (run this file)
#f = "BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt"
#N = 20534  # total number of lines
#h = 2      # header lines
# Read the first header line with sample IDs:
#id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
# Read the 2nd header line with sample type (raw, RPKM, etc.)
#header = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=FALSE, nrows=1))
# Read the rest of the file in parallel. I prefer doMC/foreach over mclapply
# for this because foreach can combine results incrementally.
#cores = 4
#registerDoMC(cores)
#block = floor((N - h)/cores)
#t1 = proc.time()
#ans = foreach(j=1:cores, .combine=rbind) %dopar%
#{
#    skip = block * (j - 1) + h
#    nrows = ifelse(j == cores, -1, block)
#    read.table(f, sep="\t", stringsAsFactors=FALSE, skip=skip, header=FALSE, nrows=nrows)
#}
#names(ans) = header
#print(proc.time() - t1)
# (All timings shown for a quad-core Athlon home PC reading from an older
# OCZ REVODRIVE3 X2 SSD drive (ca 2014).)
#   user  system elapsed
#494.549   4.134 131.232
# Compare with sequential approach:
#t1 = proc.time()
#x = read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=TRUE)
#print(proc.time() - t1)
#   user  system elapsed
#517.430   1.318 519.553
# Create a samples x gene matrix of RPKM values
#genes = ans[, 1] # gene labels
#idx   = grep("RPKM", names(ans))    # Select columns with RPKM values
#rpkm     = t(as.matrix(ans[, idx]))
# Label the coordinates by sample ID and gene:
#rownames(rpkm) = id[idx]
#colnames(rpkm) = genes
#save(rpkm, file="rpkm.RData")
# --------------------------------------------------------------------------------------------------
# Default values:
t1=40.34
t2=35.60
m1=39.10
m2=9595.10

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
load("rpkm.RData")  # rpkm expression matrix

# The data have a few columns of all zeros that need to be removed
n = ncol(rpkm)
rpkm = rpkm[, colMeans(rpkm) > 0]

threshold = 0.99     # correlation threshold
xx  = gc()           # clear
m1  = sum(gc()[,2])  # Memory in megabytes currently in use
t1  = proc.time()
tx  = tcor(rpkm, threshold, 11)
t1  = (proc.time() - t1)[3] # Time in seconds
m1  = sum(gc()[,6] - m1)    # Peak excess memory use during the test

# Compare with the fastest way we know of to compute a thresholded correlation
# matrix by brute force. This approach makes heavy use of fast multi-threaded
# BLAS routines. This approach is much faster than R's default `cor` function.
xx  = gc()
m2  = sum(gc()[,2])  # Memory in megabytes currently in use
t2 = proc.time()
mu = colMeans(rpkm)
s  = sqrt(apply(rpkm, 2, crossprod) - nrow(rpkm) * mu ^ 2) # col norms of centered matrix
cs = scale(rpkm, center=mu, scale=s)  # scale and center the data
bx = crossprod(cs)                    # full correlation matrix
diag(bx) = 0                          # ignore diagonal (all ones)
pairs = which(bx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3] # Time in seconds
m2  = sum(gc()[,6] - m2)    # Peak excess memory use during the test
} # RUN_EXAMPLES=TRUE

require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA BRCA gene expression results (878 samples x 20,542 genes, threshold=0.99)", label = "TCGA1", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

We simulated a data set with a larger number of samples by simply replicating
the rows of the gene expression matrix 27 times to obtain a matrix with 23,706
rows and 20,524 columns, noting that the simulated data have the same rank (of
at most 878) and the same expected number of vector pairs meeting the
correlation threshold as the original data. Each method again computed the same
set of  3,572 pairs of vectors meeting the correlation threshold of 0.99, but
the {\tt tcor()} function outperforms the brute force method in both compute
time and memory consumption.

<<echo=FALSE, results=tex>>=
# Default values:
t1=372.69
t2=546.06
m1=768.3
m2=9600.2

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

The following example uses calculated beta values from the Illumina Human
Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor samples
obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}. See the
supplemental code\cite{sup} for examples of efficiently converting the
downloaded sample methylation data into an R matrix.

We removed 91,563 columns with all constant or missing values from the data,
leaving an $80\times 394014$ matrix. The example computes
pairs of methylation beta value column vectors with correlation values of 0.99
or more. We use the same {\tt tcor} projection dimension of 10 and performance
measurement methodology as used in the first example.

Algorithm \ref{prune} identified 913,601 pairs of vectors with correlation
values of 0.99 or more in about three hours on our test PC. A brute force
algorithm would need to evaluate over 155 billion vector pairs; the projection
of Algorithm \ref{prune} limited the search space to only about 4 million
candidate pairs, from which the 913,601 correlated pairs were found.

We illustrate that steps 4--7 of Algorithm \ref{prune} can be computed in
parallel by comparing the above results run on our reference test PC with
results run on a modest GNU/Linux cluster at Paradigm4, Inc. The cluster
consisted of four computers connected by 10 gigabit ethernet. Each computer
contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
computer) and 64$\,$GB ECC registered RAM running at 1,600$\,$MHz running
CentOS 6.6 and 64-bit R version 3.2.2 with the reference R BLAS library.

@@

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
 # methylation data
}
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall-clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
@
From a set of almost 400 thousand vectors, Algorithm \ref{prune} finds 913,601
pairs of vectors with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm also runs well in constrained
memory settings.

@@

XXX

1. The main ideas in Algorithm \ref{prune}, although applied to correlation,
easily extend to any distance metric.

2. Anti-correlated pairs? Anyone?

@@

\begin{thebibliography}{99}

%% Related use of SVD in other methods
%\bibitem{spca}
%Bair, Eric, et al. Prediction by supervised principal components. Journal of the American Statistical Association 101.473 (2006).

%% irlba
\bibitem{irlba}
Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.

\bibitem{irlbam}
Baglama, James,  {\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}}.

%% BicatYeast data
\bibitem{bicat} Barkow, S., Bleuler, S., Prelic, A., Zimmermann, P., and E. Zitzler. BicAT: a biclustering analysis toolbox, Bioinformatics, 2006 22(10):1282-1283.

%% BROAD GDAC dashboard
\bibitem{gdac}{Broad Institute TCGA Genome Data Analysis Center (2015): Analysis-ready standardized TCGA data from Broad GDAC Firehose stddata\_\_2015\_06\_01 run. Broad Institute of MIT and Harvard. Dataset. \url{http://dx.doi.org/10.7908/C1251HBG}}

%% Applications in machine learning
\bibitem{ml1}
Hall, Mark A. Correlation-based feature selection for machine learning. Diss. The University of Waikato, 1999.

%% R's biclust pacakge
\bibitem{biclust}
Kaiser, Sebastian, Santamaria, Rodrigo, Khamiakova, Tatsiana, Sill, Martin,
Theron, Roberto, Quintales, Luis, Leisch, Friederich, and De Troyer, Ewoud, 
\url{https://cran.r-project.org/web/packages/biclust}.

%% IRLBA implementations
\bibitem{irlbap}
Kane, Michael and Lewis, Bryan,  \url{https://github.com/bwlewis/irlbpy}.

%% doredis
\bibitem{doredis}
Lewis, Bryan. The doRedis package for R. \url{https://cran.r-project.org/web/packages/doRedis}.

%% IRLBA implementations and supplemental code
\bibitem{irlbar}
Lewis, Bryan. The irlba package for R.  \url{https://cran.r-project.org/web/packages/irlba}.

\bibitem{sup}
Lewis, Bryan. The tcor package for R.  \url{https://github.com/bwlewis/tcor}.

%% Approximate correlation
\bibitem{timeseries}
Mueen, Abdullah, Suman Nath, and Jie Liu. Fast approximate correlation for massive time-series data. Proceedings of the 2010 ACM SIGMOD International Conference on Management of data. ACM, 2010.

%% Applications in genomics
\bibitem{genomics}
Song, Lin, Peter Langfelder, and Steve Horvath. Comparison of co-expression measures: mutual information, correlation, and model based indices. BMC bioinformatics 13.1 (2012): 328.

%% foreach
\bibitem{foreach}
Weston, Steve. The foreach package for R. \url{https://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf}.

%% SVD similarity -- extremely similar to the approach in this paper!
%% Applied to images
\bibitem{svd-similarity}
Wu, Daniel, et al. Efficient retrieval for browsing large image databases. Proceedings of the fifth international conference on Information and knowledge management. ACM, 1996.

%% Other correlation pruning methods
\bibitem{prune1}
Xiong, Hui, et al. TAPER: A two-step approach for all-strong-pairs correlation query in large databases. Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
\bibitem{prune2}
Xiong, Hui, Mark Brodie, and Sheng Ma. Top-cop: Mining top-k strongly correlated pairs in large databases. Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.

%% Applications in machine learning
\bibitem{ml2}
Yu, Lei, and Huan Liu. Feature selection for high-dimensional data: A fast correlation-based filter solution. ICML. Vol. 3. 2003.

%% Other correlation pruning methods
\bibitem{prune3}
Zhang, Jian, and Joan Feigenbaum. Finding highly correlated pairs efficiently with powerful pruning. Proceedings of the 15th ACM international conference on Information and knowledge management. ACM, 2006.

\end{thebibliography}
\end{document}

>>>>

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
\label{sumlemma}
Let $a_i$ and $a_j$
be any two columns of the matrix
$A$, $1\le i,j\le n$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$--that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Let $t$ be a given correlation threshold, $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders the entries
of $v_1$ in increasing order.  For example, Figure \ref{fig1} displays the
ordered entries of $Pv_1$ for the small ``EisenYeast'' example (where
$A\in\mathcal{R}^{80 \times 6221}$) from Section \ref{examples}.  The lines in
the plot illustrate the interval associated with the correlation threshold
$t=0.99$ placed at an arbitrary vertical axis location.

\begin{figure}[!ht]
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(EisenYeast)
A = t(EisenYeast)
A = sweep(A, 2, apply(A,2,mean), FUN=`-`)
A = sweep(A, 2, sqrt(apply(A,2,crossprod)), FUN=`/`)
s = svd(A)
i = order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(0, sqrt(2*(1-t))/s$d[1]), col=2, lwd=1)
@
\caption{
Example plot of the ordered entries of $v_1$ for the EisenYeast example from
Section \ref{examples}. The red lines illustrate the interval corresponding
threshold $t=0.99$ placed at an arbitrary location on the vertical axis. Points
farther apart than this interval correspond to column vectors that do not meet
the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

@@

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 96\% of
adjacent vectors fall below the correlation threshold of $0.99$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout.  In general, using $p\le k$ terms to prune pairs of vectors
below the correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

A full/na\"ive thresholded correlation computation requires $O(n^2 m)$ flops.
Let $p$ be the selected IRLBA dimension. Then the proposed method requires
$O(n p \ell)$ flops, where $\ell$ is the longest run of ordered entries in
$v_1$ that meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
plus the flops required by IRLBA and the post-processing step 7 (also $O(np)$).

@@

We evaluated the algorithm using two public gene expression and methylation
datasets.  Most tests were performed on a desktop PC with a single AMD
A10-7850K 3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered
RAM running the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R
version 3.1.2 using double-precision arithmetic and the OpenBLAS library
version 0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}. The last experiment includes a test run on a Linux
cluster as described in the text.

@@

\subsection{Small EisenYeast example}

The first example is a small proof of concept that provides the data for Figure
\ref{fig1} and establishes our timing and memory use measurement methodology.
Wall-clock times are reported in seconds and peak memory use in excess of
memory required to store the input data matrix is reported in megabytes.

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} R package\cite{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as a $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(irlba)
source("tcor.R")  # XXX Replace later with tcor pacakge XXX XXX
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

@@

# Fast brute force full correlation:
m2  = sum(gc()[,2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "EisenYeast results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

Performance benefits from the {\tt tcor()} function become apparent with more
data.  The following example uses calculated beta values from the Illumina
Human Methylation450 BeadChip of 80 Adenoid Cystic Carcinoma (ACC) tumor
samples obtained from the Broad GDAC Cancer Genome Atlas dashboard\cite{gdac}.
See the supplemental code\cite{sup} for examples of efficiently converting the
downloaded sample methylation data into an R matrix. We chose these data for
their public availability as a general example of computing thresholded
correlation among sets of large numbers of vectors.

We removed 91,563 columns with all constant or missing values from the data,
leaving an $80\times 394014$ matrix. The example computes pairs of methylation
beta value column vectors with correlation values of 0.99 or more. We used a
{\tt tcor} projection dimension of 10 and the same performance measurement
methodology used in the first example. Despite its low rank and modest data
size, the problem is too large to compute using a na\"ive brute force
correlation method.

Algorithm \ref{prune} identified 913,601 pairs of vectors with correlation
values of 0.99 or more in about three hours on our test PC. A brute force
algorithm would need to evaluate the threshold condition for more than 155
billion vector pairs; the projection of Algorithm \ref{prune} limited the
search space to only about 4 million candidate pairs, from which the 913,601
correlated pairs were found.

We illustrate that steps 4--7 of Algorithm \ref{prune} can be computed in
parallel by comparing the above results run on our reference test PC with
results run on a modest GNU/Linux cluster at Paradigm4, Inc. The cluster
consisted of four computers connected by 10 gigabit ethernet. Each computer
contained two Intel Xeon E5-2650 2$\,$GHz CPUs (16 physical CPU cores per
computer) and 64$\,$GB/1,600$\,$MHz ECC registered RAM running CentOS 6.6 and
64-bit R version 3.2.2 with the reference R BLAS library.

@@

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
 # methylation data
}
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
@
From a set of almost 400 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm also runs, although slower, in
the constrained memory setting of a desktop PC.

@@

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation. Lemma
\ref{sumlemma} decomposes Euclidean distance between two vectors in terms of
singular basis vectors and Algorithm \ref{prune} can be applied directly to the
problem of finding all pairs of vectors that meet a thresholded Euclidean
distance with small changes in steps 3, 4, and 7.  And, by equivalence of
norms, thresholded Euclidean distance can be extended to other distance
metrices similarly. The reference R implementation\cite{sup} includes functions
for general thresholded distance computation.

@@

\bibliography{bibliography}{}
\bibliographystyle{plain}

\end{document}
